Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-09-20 intro.tex

\section{Introduction to quantum chromodynamics}\label{sec:qcd}
    \subsection{The QCD Lagrangian}\label{sec:L_qcd}
    Quantum chronodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{c}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{s}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{s} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, 
    which is an $N_{c}^{2}-1$ dimensional representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The generators are normalised by convention to be
    \begin{equation}\label{eqn:generator_normalisation}
        \mathrm{Tr}(T^{a}T^{b}) = \delta^{ab}T_{R} \, , \quad T_{R} = \dfrac{1}{2} \, .
    \end{equation}
    This normalisation sets the values of the Casimirs
    of the group as
    \begin{align}\label{eqn:casimirs}
        \begin{split}
            T^{a}_{ij}T^{a}_{jk} &= \delta{ik}C_{F} \, , \quad C_{F} = \dfrac{N_{c}^2-1}{2N_{c}} \, , \\
            f^{abc}f^{abd} &= \delta^{cd}C_{A} \, , \quad C_{A} = N_{c} \, ,
        \end{split}
    \end{align}
    where $C_{F}=4/3$ and $C_{A}=3$ for QCD.
    They are collectively referred to as colour factors.

    The gauge coupling of the group $g_{s}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{s} = \dfrac{g_{s}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process. This
    is due to the process of renormalisation which will be discussed
    in section ...
    The implications of this is that at collider
    experiments where collisions are extremely high energy,
    the strong coupling constant becomes small, a
    propery known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{s}$.

@@

    This heuristic argument motivates the form of
    the factorisation equation, where the production of
    particles from a hadronic collision, the hadronic
    cross-section, can be written as
    \begin{equation}\label{eqn:hadronic_cs}
        \sigma_{AB \rightarrow n} = \sum_{a, b}\int_{0}^{1} \mathrm{d}x_{a}\mathrm{d}x_{b} \; f_{a/A}(x_{a},\mu_{F})f_{b/B}(x_{b}, \mu_{F}) \; \hat{\sigma}_{ab \rightarrow n}(Q, \mu_{F}, \mu_{R}) + \mathcal{O}\left(\dfrac{\Lambda_{\mathrm{QCD}}}{Q}\right)\, ,
    \end{equation}
    where $a$ is the parton from hadron $A$,
    and $b$ is the parton from hadron $B$.
    The sum over initial-state partons $a$ and $b$
    indicates the sum over flavours, which depends
    on the hadron composition in general.
    This factorisation is not exact as indicated by the
    correction term inversely proportional to characteristic
    hard scale $Q$. However, for high energy collisions, where
    $Q \gg \Lambda_{\mathrm{QCD}}$, this term and higher order
    terms are suppresed.
    In fact, the factorisation equation has only been
    proven for a few specific cases (DIS etc, cite here).

    $f_{a/h}(x, \mu)$ are parton distribution functions
    (PDFs) which depend on the momenta fractions $x$ parton $a$
    has with respect to its parent hadron $h$, at a scale $\mu$,
    usually taken to be the factorisation scale $\mu_{F}$,
    the interface between soft and hard physics.
    The interpretation of PDFs at leading order are
    as probability distributions. $f_{a/h}(x, \mu)$ is
    the probability of finding parton $a$ within $h$
    carrying a momentum fraction $x$ at the energy scale $\mu$.
    PDFs are non-perturbative objects that encapsulate
    the soft effects of the scattering which occur below energy
    $\mu_{F}$. They are non-perturbative because
    they are determined by fitting to experimental data,
    instead of calculated in perturbation theory.
    Multiple groups release PDF fits (cite groups here).
    In practice, these PDF fits
    are accessed through the LHAPDF interface (cite here).

    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations (cite here)
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ba}(\dfrac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$ carrying
    a momentum fraction $z$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops (cite here). At leading order (cite here) they are
    \begin{align}\label{eqn:AP_kernels}
        P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \\
        P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \\
        P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \\
        P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z)
    \end{align}
    where the `$+$'-prescription has been used to regulate divergences
    at $z = 1$.
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, ,
    \end{equation}
    where the function $f(z)$ cancels the divergence if it is
    sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a fraction $x$ of the parent parton.

>>>>

\section{Introduction to quantum chromodynamics}\label{sec:qcd}
    \subsection{The QCD Lagrangian}\label{sec:L_qcd}
    Quantum chromodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{c}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{s}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{s} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, 
    which is an $N_{c}^{2}-1$ dimensional representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The generators are normalised by convention to be
    \begin{equation}\label{eqn:generator_normalisation}
        \mathrm{Tr}(T^{a}T^{b}) = \delta^{ab}T_{R} \, , \quad T_{R} = \dfrac{1}{2} \, .
    \end{equation}
    This normalisation sets the values of the Casimirs
    of the group as
    \begin{align}\label{eqn:casimirs}
        \begin{split}
            T^{a}_{ij}T^{a}_{jk} &= \delta{ik}C_{F} \, , \quad C_{F} = \dfrac{N_{c}^2-1}{2N_{c}} \, , \\
            f^{abc}f^{abd} &= \delta^{cd}C_{A} \, , \quad C_{A} = N_{c} \, ,
        \end{split}
    \end{align}
    where $C_{F}=4/3$ and $C_{A}=3$ for QCD.
    They are collectively referred to as colour factors.

    The gauge coupling of the group $g_{s}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{s} = \dfrac{g_{s}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process. This
    is due to the process of renormalisation which will be discussed
    in section ...
    The implications of this is that at collider
    experiments where collisions are extremely high energy,
    the strong coupling constant becomes small, a
    property known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{s}$.

@@

    This heuristic argument motivates the form of
    the factorisation equation, where the production of
    particles from a hadronic collision, the hadronic
    cross-section, can be written as
    \begin{equation}\label{eqn:hadronic_cs}
        \sigma_{AB \rightarrow n} = \sum_{a, b}\int_{0}^{1} \mathrm{d}x_{a}\mathrm{d}x_{b} \; f_{a/A}(x_{a},\mu_{F})f_{b/B}(x_{b}, \mu_{F}) \; \hat{\sigma}_{ab \rightarrow n}(Q, \mu_{F}, \mu_{R}) + \mathcal{O}\left(\dfrac{\Lambda_{\mathrm{QCD}}}{Q}\right)\, ,
    \end{equation}
    where $a$ is the parton from hadron $A$,
    and $b$ is the parton from hadron $B$.
    The sum over initial-state partons $a$ and $b$
    indicates the sum over flavours, which depends
    on the hadron composition in general.
    This factorisation is not exact as indicated by the
    correction term inversely proportional to characteristic
    hard scale $Q$. However, for high energy collisions, where
    $Q \gg \Lambda_{\mathrm{QCD}}$, this term and higher order
    terms are suppressed.
    In fact, the factorisation equation has only been
    proven for a few specific cases \cite{Collins:2011zzd,Collins:1987pm,Collins:1989gx,Amoroso:2022eow}.

    $f_{a/h}(x, \mu)$ are parton distribution functions
    (PDFs) which depend on the momenta fractions $x$ parton $a$
    has with respect to its parent hadron $h$, at a scale $\mu$,
    usually taken to be the factorisation scale $\mu_{F}$,
    the interface between soft and hard physics.
    The interpretation of PDFs at leading order are
    as probability distributions. $f_{a/h}(x, \mu)$ is
    the probability of finding parton $a$ within $h$
    carrying a momentum fraction $x$ at the energy scale $\mu$.
    PDFs are non-perturbative objects that encapsulate
    the soft effects of the scattering which occur below energy
    $\mu_{F}$. They are non-perturbative because
    they are determined by fitting to experimental data,
    instead of calculated in perturbation theory.

    For a review of PDFs and how they are determined
    see Ref. \cite{Ethier:2020way,Jimenez-Delgado:2013sma}.
    In practice, these PDF fits are accessed through the
    LHAPDF interface \cite{Buckley:2014ana} which provides
    PDF sets from multiple working groups \cite{PDF4LHCWorkingGroup:2022cjn,Hou:2019efy,Bailey:2020ooq,NNPDF:2021njg}.
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations \cite{Dokshitzer:1977sg,Gribov:1972ri,Lipatov:1974qm,Altarelli:1977zs}.
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ba}(\dfrac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$ carrying
    a momentum fraction $z$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops (cite here). At leading order (cite here) they are
    \begin{align}\label{eqn:AP_kernels}
        P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \\
        P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \\
        P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \\
        P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z)
    \end{align}
    where the `$+$'-prescription has been used to regulate divergences
    at $z = 1$.
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, ,
    \end{equation}
    where the function $f(z)$ cancels the divergence if it is
    sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a fraction $x$ of the parent parton.

### 2022-09-20 qcd.tex

\section{Factorisation of matrix elements}\label{sec:me_factorisation}

>>>>

\section{Factorisation of matrix elements}\label{sec:me_factorisation}
    In the previous section we examined the IR divergences
    that can arise in matrix elements, either explicitly
    through loop diagrams, or implicitly when carrying
    out the phase-space integral over the external state
    momenta. In this section we will discuss how these IR
    singularities are universal and can be factorised out
    of the matrix element. This property is exploited
    extensively in Sections ... where the research of this
    thesis is presented.
